---
title: "Project"
output: html_document
date: "2025-05-15"
---

```{r}
library(readr,tidytuesdayR)
```

```         
```

```         
```

```{r}
historical_spending <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-02-13/historical_spending.csv', show_col_types = FALSE)
gifts_age <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-02-13/gifts_age.csv', show_col_types = FALSE)
gifts_gender <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-02-13/gifts_gender.csv', show_col_types = FALSE)

```

```{r}
historical_spending
```

```{r}
gifts_age
```

```{r}
gifts_gender
```

# EDA

```{r}
library(tidyverse)
library(lubridate)
library(scales)
library(skimr)
library(patchwork)
library(ggrepel)
library(dplyr)
library(tidyr)
```

```{r}
glimpse(historical_spending)
```

```{r}
skim(historical_spending)
```

```{r}
skim(gifts_age)
```

```{r}
skim(gifts_gender)
```

```{r}
sum(is.na(gifts_age)) 
```

```{r}
sum(is.na(historical_spending))
```

```{r}
gift_gender_wide <- gifts_gender %>% 
  pivot_wider(
    names_from = Gender,
    values_from = Gender
  )
gift_gender_wide
```

```{r}
gifts_gender
```

# Dataset's extention

## historical_spending

### Compute total average spending per year (AnnualSpending_tot)

```{r}
historical_spending_ext <- historical_spending %>% 
  mutate(AnnualSpending_tot = rowSums(.[,4:ncol(.)]))

historical_spending_ext
```

### Compute change in PerPerson spending year-over-year (SpendingChange)

```{r}
historical_spending_ext <- historical_spending_ext %>%
  arrange(Year) %>% 
  mutate(SpendingChange = case_when(
           row_number() == 1 ~ "0%",
           !is.na(PerPerson) & !is.na(lag(PerPerson)) ~ {
             change <- (PerPerson - lag(PerPerson)) / lag(PerPerson)
             ifelse(change > 0, paste0("+", round(change * 100, 2), "%"),
             paste0(round(change * 100,2), "%"))
            },
           TRUE ~ NA_character_
         )
         )

historical_spending_ext
```

### Find category with the highest spending over the year (HighestSpending)

```{r}
historical_spending_ext <- historical_spending_ext %>% 
  mutate(
    SpendingCols = across(Candy:GiftCards),
    HighestSpending = names(SpendingCols)[max.col(SpendingCols, ties.method = "first")],
    SpendingCols = NULL
  )
historical_spending_ext
```

## gifts_age

### Convert Age to categorical (Age_cat)

```{r}
gifts_age_cat <- gifts_age %>% 
  mutate(
    Age_cat = case_when(
      Age == "18-24" ~ 1,
      Age == "25-34" ~ 2,
      Age == "35-44" ~ 3,
      Age == "45-54" ~ 4,
      Age == "55-64" ~ 5,
      Age == "65+" ~ 6,
      TRUE ~ NA_integer_
    ),
    .after = Age
  )
print(gifts_age_cat)
```

### Find category with the highest spending by each group (HighestSpending)

```{r}
gifts_age_ext <- gifts_age_cat %>% 
  mutate(
    SpendingCols = across(Candy:GiftCards),
    HighestSpending = names(SpendingCols)[max.col(SpendingCols, ties.method = "first")],
    SpendingCols = NULL
  )
gifts_age_ext
```

### Compute Affinity index between Age and average spending on each category (Affinity\_ ). An Affinity index \>100 indicate that the target group is more likely to purchase some products compared to the average across all age groups

```{r}
overall_cat_relevance <- gifts_age_ext %>% 
  summarise(across(Candy:GiftCards, ~ mean(.x, na.rm=TRUE))) %>% 
  pivot_longer(cols=everything(), names_to = "Category", values_to = "OverallRelevance")

gifts_age_affinity <- gifts_age_ext %>% 
  pivot_longer(cols = Candy:GiftCards, names_to = "Category", values_to = "AgeGroupRelevance") %>% 
  left_join(overall_cat_relevance, by="Category") %>% 
  mutate(
    AffinityIndex = ifelse(
      OverallRelevance != 0,
      (AgeGroupRelevance / OverallRelevance) * 100,
      NA_real_
    )
  ) %>% 
  select(Age, Category, AffinityIndex) %>% 
  pivot_wider(names_from =Category, values_from = AffinityIndex, names_prefix = "Affinity_")

gifts_age_ext_wide <- left_join(gifts_age_ext, gifts_age_affinity, by = "Age")

gifts_age_ext_wide
```

## gifts_gender

### Find category with the highest spending by each gender (HighestSpending)

```{r}
gifts_gender_ext <- gifts_gender %>% 
  mutate(
    SpendingCols = across(Candy:GiftCards),
    HighestSpending = names(SpendingCols)[max.col(SpendingCols, ties.method = "first")],
    SpendingCols = NULL
  )
gifts_gender_ext
```

### Compute Affinity index between Gender and average spending on each category (Affinity\_ ). An Affinity index \>100 indicate that the target group is more likely to purchase some products compared to the average across all genders

```{r}
overall_gender_cat_relevance <- gifts_gender_ext %>% 
  summarise(across(Candy:GiftCards, ~ mean(.x, na.rm=TRUE))) %>% 
  pivot_longer(cols=everything(), names_to = "Category", values_to = "OverallRelevance")

gifts_gender_affinity <- gifts_gender_ext %>% 
  pivot_longer(cols = Candy:GiftCards, names_to = "Category", values_to = "GenderRelevance") %>% 
  left_join(overall_gender_cat_relevance, by="Category") %>% 
  mutate(
    AffinityIndex = ifelse(
      OverallRelevance != 0,
      (GenderRelevance / OverallRelevance) * 100,
      NA_real_
    )
  ) %>% 
  select(Gender, Category, AffinityIndex) %>% 
  pivot_wider(names_from =Category, values_from = AffinityIndex, names_prefix = "Affinity_")

gifts_gender_ext_wide <- left_join(gifts_gender_ext, gifts_gender_affinity, by = "Gender")

gifts_gender_ext_wide
```

# Plots

### Affinity Index vs Age Group

```{r}
age_ordered <- gifts_age_ext_wide %>% 
  distinct(Age, Age_cat) %>% 
  arrange(Age_cat) %>% 
  pull(Age)

gifts_age_affinity_plot <- gifts_age_ext_wide %>% 
  select(Age, starts_with("Affinity_")) %>% 
  pivot_longer(
    cols = starts_with("Affinity_"),
    names_to = "Category",
    values_to = "AffinityIndex"
  ) %>% 
  mutate(
    Category = gsub("Affinity_","",Category),
    Age = factor(Age, levels = age_ordered)
  )

age_trend <- gifts_age_affinity_plot %>% 
  group_by(Age) %>% 
  summarise(AverageAffinityIndex = mean(AffinityIndex, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(Age=factor(Age, levels = age_ordered))

ggplot(gifts_age_affinity_plot, aes(x=Age, y=AffinityIndex)) + 
  geom_col(aes(fill=Category),position = "dodge") + 
  geom_hline(yintercept = 100, linetype = "dashed", color="grey50",linewidth=0.8) +
  geom_line(data = age_trend, aes(x=Age, y=AverageAffinityIndex, group=1), 
            color ="pink", linetype="solid", linewidth=1.2) +
  geom_point(data=age_trend, aes(x=Age, y=AverageAffinityIndex, group=1),
             color="pink", size=3, shape=3, shape=21, fill="black") +
  theme_minimal() +
  labs(
    title = "Affinity Index vs Age Group (By Gift Categories)",
    y = "Affinity Index",
    x = "Age Group"
  ) + 
  scale_fill_brewer(palette = "Set3") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

#### The above plot visualizes the affinity indexes for each category of gifts and age group. We decided to include this plot as it shows the most popular categories of gifts as well as some trends. An Affinity Index above 100 indicates that an age group is more likely to buy that category of gifts compared to the average (100), while below 100 indicates less likelihood.

#### Interpretation:

#### - Among the 18- 44 years audience, the most popular categories of gifts are Clothing and Jewelry. For 45+ audience, the most popular category of gifts is Greeting Cards.

#### - We may observe the overall trend (pink line) that the percent of people who buy gifts decreases over time.

## Affinity Index vs Gender

```{r}
gifts_gender_affinity_plot <- gifts_gender_ext_wide %>% 
  select(Gender, starts_with("Affinity_")) %>% 
  pivot_longer(
    cols= starts_with("Affinity_"),
    names_to = "Category",
    values_to = "AffinityIndex"
  ) %>% 
  mutate(
    Category = gsub("Affinity_", "", Category),
    Gender = factor(Gender, levels=c("Men", "Women"))
  )

ggplot(gifts_gender_affinity_plot, aes(x=Category, y=AffinityIndex, fill=Gender)) + 
  geom_col(position="dodge") + 
  geom_hline(yintercept = 100, linetype="dashed", color="grey50", linewidth=0.8) +   theme_minimal() + 
  labs(
    title="Affinity Index vs Gift Categories (By Gender)",
    y = "Affinity Index",
    x = "Gift Category"
  ) + 
  scale_fill_manual(values=c("Men"="steelblue", "Women" = "lightcoral")) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### The above plot visualizes the affinity indexes for each category of gifts and gender. We decided to include this plot as it shows the most popular categories of gifts. An Affinity Index above 100 indicates that a gender is more likely to buy that category of gifts compared to the average (100), while below 100 indicates less likelihood.

#### Interpretation:

#### - On average, women are more likely to buy gifts with the exception of Flowers and Jewelry.

### Change of Total Spending over time

```{r}
inflation_data <- tibble(
  Year = c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025),
  Amount = c(1, 1.03, 1.05, 1.07, 1.09, 1.09, 1.1, 1.12, 1.15, 1.17, 1.19, 1.24, 1.34, 1.4, 1.44, 1.47),
  InflationRate = c(0.02, 0.03, 0.02, 0.01, 0.02, 0, 0.01, 0.02, 0.02, 0.02, 0.01, 0.05, 0.08, 0.04, 0.03, 0.02)
) %>%
  rename(CPI_Index = Amount)

historical_spending_w_inflation <- historical_spending_ext %>%
  left_join(inflation_data %>% select(Year, CPI_Index), by = "Year")

base_year_cpi_index <- inflation_data %>% filter(Year == 2022) %>% pull(CPI_Index)

historical_spending_adjusted <- historical_spending_w_inflation %>%
  mutate(
    Real_AnnualSpending_tot = AnnualSpending_tot * (base_year_cpi_index / CPI_Index)
  ) %>%
  select(Year, AnnualSpending_tot, Real_AnnualSpending_tot) %>%
  pivot_longer(
    cols = c(AnnualSpending_tot, Real_AnnualSpending_tot),
    names_to = "SpendingType",
    values_to = "SpendingValue"
  )

ggplot(historical_spending_adjusted, aes(x = Year, y = SpendingValue, color = SpendingType)) +
  geom_line(size = 1.2) +
  geom_point(size = 3, shape = 19) +
  scale_x_continuous(breaks = unique(historical_spending_adjusted$Year)) +
  scale_color_manual(
    values = c("AnnualSpending_tot" = "purple", "Real_AnnualSpending_tot" = "darkgreen"),
    labels = c("AnnualSpending_tot" = "Nominal Spending", "Real_AnnualSpending_tot" = paste0("Real Spending (2022 $)"))
  ) +
  theme_minimal() +
  labs(
    title = "Total Spending Over Time",
    y = "Total Annual Spending ($)",
    x = "Year",
    color = "Spending Type"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

#### The above plot represents the change of spending over time. We took into account inflation of dollar and added an adjusted line (green) to visualize the actual spending volume change. The inflation data has been take from the open source: <https://www.in2013dollars.com/us/inflation/2010?endYear=2022&amount=1>

#### Interpretation:

#### - Both Nominal and Real Spending lines indicate that there's an overall increase in the actual purchasing power of money spent over time.

### Average Spending per Gift category

```{r}
historical_spending_ext_long <- historical_spending_ext %>% 
  select(1, 4:(ncol(historical_spending_ext) - 3)) %>% 
  pivot_longer(
    cols = -Year,
    names_to = "Category",
    values_to = "AverageSpending"
  )

historical_spending_ext_long
ggplot(historical_spending_ext_long, aes(x=Year, y=AverageSpending, fill=Category)) + 
  geom_area(alpha=0.8, position = "stack") + 
  scale_x_continuous(breaks=unique(historical_spending_ext_long$Year)) + 
  theme_minimal() + 
  labs(
    title = "Average Spending per Gift category over Time",
    y = "Average Spending",
    x = "Year",
    fill = "Gift Category"
  ) + scale_fill_brewer(palette = "Paired") + 
  theme(plot.title = element_text(hjust=0.5))
```

#### Stacked area plot appears to be a perfect tool to visualize the proportion of spending on each category of gifts and the overall annual change.

#### Interpretation:

#### - Jewelry and Evening Out represent the largest portions of total spending over the observed period

#### - Spending on all categories showed limited growth, even experiencing declines, during the 2015-2019 period.

## Distribution of Spending Across Each Gift Category Over All Years

```{r}
historical_spending_long_categories <- historical_spending_ext %>%
  select(Year, Candy:GiftCards) %>%
  pivot_longer(cols = -Year, names_to = "Category", values_to = "Spending")

ggplot(historical_spending_long_categories, aes(x = Category, y = Spending, fill = Category)) +
  geom_boxplot() +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = "Distribution of Spending Across Gift Categories Over All Years",
    x = "Gift Category",
    y = "Spending ($)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

#### This box plot displays the range, median, and variability of actual dollar spending for each specific gift category over all years, serving to highlight which categories consistently involve higher or lower expenditures and their spread.

## Gift Category Spending Distribution by Age

```{r}

gifts_age_percentages_long_categories <- gifts_age_ext %>%
  select(Age, Age_cat, Candy:GiftCards) %>% 
  pivot_longer(
    cols = Candy:GiftCards, 
    names_to = "Category",
    values_to = "PercentageOfPeople"
  ) %>%

  mutate(Age = factor(Age, levels = unique(gifts_age_ext$Age[order(gifts_age_ext$Age_cat)])))

ggplot(gifts_age_percentages_long_categories, aes(x = Category, y = PercentageOfPeople, fill = Category)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "Distribution of Percentage of People Spending Across Gift Categories by Age Group",
    x = "Gift Category",
    y = "Percentage of People (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") 
```

#### This box plot visualizes the statistical distribution of the percentage of people who purchase different gift categories, allowing for a clear comparison of varying buying prevalences across those categories within specific age groups.

## Percentage of Celebrating by Age Group

```{r}
ggplot(gifts_age_ext_wide, aes(x=Age, y=SpendingCelebrating)) + 
  geom_col(fill="steelblue") +
  geom_text(aes(label=paste0(round(SpendingCelebrating, 1), "%")),
            vjust= -0.5,
            color="black",
            size=3) + 
  theme_minimal() +
  labs(
    title = "Percentage of People Celebrating by Age",
    y = "Percent of Celebrating",
    x = "Age Group"
  ) + 
  theme(axis.title.x = element_text(angle = 45, hjust=1))

```

#### This plot compares how many individuals from every age group are celebrating Saint Valentine's Day

#### Interpretation:

#### - The 18-24 appears to have the highest percentages of people celebrating, there's a strictly inversed dependency with the age

```{r}
write.csv(gifts_gender_ext_wide, 'gifts_gender_processed.csv')
```

```{r}
write.csv(gifts_age_ext_wide, 'gifts_age_processed.csv')
```
