---
title: "Project"
output: html_document
date: "2025-05-15"
---

```{r}
library(readr,tidytuesdayR)
```

```         
```

```         
```

```{r}
historical_spending <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-02-13/historical_spending.csv', show_col_types = FALSE)
gifts_age <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-02-13/gifts_age.csv', show_col_types = FALSE)
gifts_gender <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-02-13/gifts_gender.csv', show_col_types = FALSE)

```

```{r}
historical_spending
```

```{r}
gifts_age
```

```{r}
gifts_gender
```

# EDA

```{r}
library(tidyverse)
library(lubridate)
library(scales)
library(skimr)
library(patchwork)
library(ggrepel)
library(dplyr)
library(tidyr)
```

```{r}
glimpse(historical_spending)
```

```{r}
skim(historical_spending)
```

```{r}
skim(gifts_age)
```

```{r}
skim(gifts_gender)
```

```{r}
sum(is.na(gifts_age)) 
```

```{r}
sum(is.na(historical_spending))
```

```{r}
gift_gender_wide <- gifts_gender %>% 
  pivot_wider(
    names_from = Gender,
    values_from = Gender
  )
gift_gender_wide
```

```{r}
gifts_gender
```

## Dataset's extention

## historical_spending

### Compute total average spending per year (AnnualSpending_tot)

```{r}
historical_spending_ext <- historical_spending %>% 
  mutate(AnnualSpending_tot = rowSums(.[,4:ncol(.)]))

historical_spending_ext
```

### Compute change in PerPerson spending year-over-year (SpendingChange)

```{r}
historical_spending_ext <- historical_spending_ext %>%
  arrange(Year) %>% 
  mutate(SpendingChange = case_when(
           row_number() == 1 ~ "0%",
           !is.na(PerPerson) & !is.na(lag(PerPerson)) ~ {
             change <- (PerPerson - lag(PerPerson)) / lag(PerPerson)
             ifelse(change > 0, paste0("+", round(change * 100, 2), "%"),
             paste0(round(change * 100,2), "%"))
            },
           TRUE ~ NA_character_
         )
         )

historical_spending_ext
```

### Find category with the highest spending over the year (HighestSpending)

```{r}
historical_spending_ext <- historical_spending_ext %>% 
  mutate(
    SpendingCols = across(Candy:GiftCards),
    HighestSpending = names(SpendingCols)[max.col(SpendingCols, ties.method = "first")],
    SpendingCols = NULL
  )
historical_spending_ext
```

## gifts_age

### Convert Age to categorical (Age_cat)

```{r}
gifts_age <- gifts_age %>% 
  mutate(
    Age_cat = case_when(
      Age == "18-24" ~ 1,
      Age == "25-34" ~ 2,
      Age == "35-44" ~ 3,
      Age == "45-54" ~ 4,
      Age == "55-64" ~ 5,
      Age == "65+" ~ 6,
      TRUE ~ NA_integer_
    ),
    .after = Age
  )
print(gifts_age)
```

### Compute total gift spending by age (GiftSpending_tot)

```{r}
gifts_age_ext <- gifts_age %>% 
  mutate(GiftSpending_tot = rowSums(.[,4:ncol(.)]))

gifts_age_ext
```

### Compute Affinity index between Age and average spending on each category (Affinity\_ )

```{r}
spending_tot <- gifts_age %>% 
  summarise(across(Candy:GiftCards, \(x) sum(x)))


gifts_age_affinity <- gifts_age_ext %>% 
  pivot_longer(cols=Candy:GiftCards, names_to = "Category", values_to = "CategorySpending") %>% 
  group_by(Age,Category) %>% 
  summarise(
    AgeTotalSpending = sum(CategorySpending),
    AffinityIndex = (AgeTotalSpending / first(SpendingCelebrating)) / 
      (first(spending_tot[[Category]]) / first(gifts_age$SpendingCelebrating))
  ) %>% 
  pivot_wider(names_from = Category, values_from = AffinityIndex, names_prefix = "Affinity_")
gifts_age_ext_wide <- left_join(gifts_age_ext, gifts_age_affinity, by="Age")

gifts_age_ext_wide
   
```

### Top 3 Affinity index values 

```{r}
top_affinities <- gifts_age_ext_wide %>% 
  pivot_longer(cols=starts_with("Affinity_"), names_to="Category", values_to = "Affinity") %>%
  slice_max(Affinity, n=3,with_ties = FALSE) %>% 
  arrange(Age, Category, desc(Affinity)) %>% 
  select(Age, Category, Affinity) %>% 
  arrange(desc(Affinity))
print(top_affinities)
```

## gifts_gender

### Compute total gift spending by gender (GiftSpending_tot)

```{r}
gifts_gender_ext <- gifts_gender %>% 
  mutate(GiftSpending_tot = rowSums(.[,3:ncol(.)]))

gifts_gender_ext
```

```{r}
### Compare gift spending by gender
gifts_gender_long <- gifts_gender_ext %>%
  pivot_longer(cols = -Gender, names_to = "Category", values_to = "Spending")

ggplot(gifts_gender_long, aes(x = Category, y = Spending, fill = Gender)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(
    title = "Gift Spending by Category and Gender",
    y = "Spending",
    x = "Gift Category"
  ) +
  scale_fill_brewer(palette = "Set1")
```
