---
title: "R_Project"
output: html_document
date: "2025-06-30"
prepared by: Tsimafei Nestsiarau, Viktoria Melnyk
---

# Part I

```{r}
library(readr,tidytuesdayR)
```

```{r}
historical_spending <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-02-13/historical_spending.csv', show_col_types = FALSE)
gifts_age <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-02-13/gifts_age.csv', show_col_types = FALSE)
gifts_gender <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-02-13/gifts_gender.csv', show_col_types = FALSE)

```

```{r}
historical_spending
```

```{r}
gifts_age
```

```{r}
gifts_gender
```

# EDA

```{r}
library(tidyverse)
library(lubridate)
library(scales)
library(skimr)
library(patchwork)
library(ggrepel)
library(dplyr)
library(tidyr)
```

```{r}
glimpse(historical_spending)
```

```{r}
skim(historical_spending)
```

```{r}
skim(gifts_age)
```

```{r}
skim(gifts_gender)
```

```{r}
sum(is.na(gifts_age)) 
```

```{r}
sum(is.na(historical_spending))
```

```{r}
gift_gender_wide <- gifts_gender %>% 
  pivot_wider(
    names_from = Gender,
    values_from = Gender
  )
gift_gender_wide
```

```{r}
gifts_gender
```

# Dataset's extention

## historical_spending

### Compute total average spending per year (AnnualSpending_tot)

```{r}
historical_spending_ext <- historical_spending %>% 
  mutate(AnnualSpending_tot = rowSums(.[,4:ncol(.)]))

historical_spending_ext
```

### Compute change in PerPerson spending year-over-year (SpendingChange)

```{r}
historical_spending_ext <- historical_spending_ext %>%
  arrange(Year) %>% 
  mutate(SpendingChange = case_when(
           row_number() == 1 ~ "0%",
           !is.na(PerPerson) & !is.na(lag(PerPerson)) ~ {
             change <- (PerPerson - lag(PerPerson)) / lag(PerPerson)
             ifelse(change > 0, paste0("+", round(change * 100, 2), "%"),
             paste0(round(change * 100,2), "%"))
            },
           TRUE ~ NA_character_
         )
         )

historical_spending_ext
```

### Find category with the highest spending over the year (HighestSpending)

```{r}
historical_spending_ext <- historical_spending_ext %>% 
  mutate(
    SpendingCols = across(Candy:GiftCards),
    HighestSpending = names(SpendingCols)[max.col(SpendingCols, ties.method = "first")],
    SpendingCols = NULL
  )
historical_spending_ext
```

## gifts_age

### Convert Age to categorical (Age_cat)

```{r}
gifts_age_cat <- gifts_age %>% 
  mutate(
    Age_cat = case_when(
      Age == "18-24" ~ 1,
      Age == "25-34" ~ 2,
      Age == "35-44" ~ 3,
      Age == "45-54" ~ 4,
      Age == "55-64" ~ 5,
      Age == "65+" ~ 6,
      TRUE ~ NA_integer_
    ),
    .after = Age
  )
print(gifts_age_cat)
```

### Find category with the highest spending by each group (HighestSpending)

```{r}
gifts_age_ext <- gifts_age_cat %>% 
  mutate(
    SpendingCols = across(Candy:GiftCards),
    HighestSpending = names(SpendingCols)[max.col(SpendingCols, ties.method = "first")],
    SpendingCols = NULL
  )
gifts_age_ext
```

### Compute Affinity index between Age and average spending on each category (Affinity\_ ). An Affinity index \>100 indicate that the target group is more likely to purchase some products compared to the average across all age groups

```{r}
overall_cat_relevance <- gifts_age_ext %>% 
  summarise(across(Candy:GiftCards, ~ mean(.x, na.rm=TRUE))) %>% 
  pivot_longer(cols=everything(), names_to = "Category", values_to = "OverallRelevance")

gifts_age_affinity <- gifts_age_ext %>% 
  pivot_longer(cols = Candy:GiftCards, names_to = "Category", values_to = "AgeGroupRelevance") %>% 
  left_join(overall_cat_relevance, by="Category") %>% 
  mutate(
    AffinityIndex = ifelse(
      OverallRelevance != 0,
      (AgeGroupRelevance / OverallRelevance) * 100,
      NA_real_
    )
  ) %>% 
  select(Age, Category, AffinityIndex) %>% 
  pivot_wider(names_from =Category, values_from = AffinityIndex, names_prefix = "Affinity_")

gifts_age_ext_wide <- left_join(gifts_age_ext, gifts_age_affinity, by = "Age")

gifts_age_ext_wide
```

## gifts_gender

### Find category with the highest spending by each gender (HighestSpending)

```{r}
gifts_gender_ext <- gifts_gender %>% 
  mutate(
    SpendingCols = across(Candy:GiftCards),
    HighestSpending = names(SpendingCols)[max.col(SpendingCols, ties.method = "first")],
    SpendingCols = NULL
  )
gifts_gender_ext
```

### Compute Affinity index between Gender and average spending on each category (Affinity\_ ). An Affinity index \>100 indicate that the target group is more likely to purchase some products compared to the average across all genders

```{r}
overall_gender_cat_relevance <- gifts_gender_ext %>% 
  summarise(across(Candy:GiftCards, ~ mean(.x, na.rm=TRUE))) %>% 
  pivot_longer(cols=everything(), names_to = "Category", values_to = "OverallRelevance")

gifts_gender_affinity <- gifts_gender_ext %>% 
  pivot_longer(cols = Candy:GiftCards, names_to = "Category", values_to = "GenderRelevance") %>% 
  left_join(overall_gender_cat_relevance, by="Category") %>% 
  mutate(
    AffinityIndex = ifelse(
      OverallRelevance != 0,
      (GenderRelevance / OverallRelevance) * 100,
      NA_real_
    )
  ) %>% 
  select(Gender, Category, AffinityIndex) %>% 
  pivot_wider(names_from =Category, values_from = AffinityIndex, names_prefix = "Affinity_")

gifts_gender_ext_wide <- left_join(gifts_gender_ext, gifts_gender_affinity, by = "Gender")

gifts_gender_ext_wide
```

# Plots

### Affinity Index vs Age Group

```{r}
age_ordered <- gifts_age_ext_wide %>% 
  distinct(Age, Age_cat) %>% 
  arrange(Age_cat) %>% 
  pull(Age)

gifts_age_affinity_plot <- gifts_age_ext_wide %>% 
  select(Age, starts_with("Affinity_")) %>% 
  pivot_longer(
    cols = starts_with("Affinity_"),
    names_to = "Category",
    values_to = "AffinityIndex"
  ) %>% 
  mutate(
    Category = gsub("Affinity_","",Category),
    Age = factor(Age, levels = age_ordered)
  )

age_trend <- gifts_age_affinity_plot %>% 
  group_by(Age) %>% 
  summarise(AverageAffinityIndex = mean(AffinityIndex, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(Age=factor(Age, levels = age_ordered))

ggplot(gifts_age_affinity_plot, aes(x=Age, y=AffinityIndex)) + 
  geom_col(aes(fill=Category),position = "dodge") + 
  geom_hline(yintercept = 100, linetype = "dashed", color="grey50",linewidth=0.8) +
  geom_line(data = age_trend, aes(x=Age, y=AverageAffinityIndex, group=1), 
            color ="pink", linetype="solid", linewidth=1.2) +
  geom_point(data=age_trend, aes(x=Age, y=AverageAffinityIndex, group=1),
             color="pink", size=3, shape=3, shape=21, fill="black") +
  theme_minimal() +
  labs(
    title = "Affinity Index vs Age Group (By Gift Categories)",
    y = "Affinity Index",
    x = "Age Group"
  ) + 
  scale_fill_brewer(palette = "Set3") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

#### The above plot visualizes the affinity indexes for each category of gifts and age group. We decided to include this plot as it shows the most popular categories of gifts as well as some trends. An Affinity Index above 100 indicates that an age group is more likely to buy that category of gifts compared to the average (100), while below 100 indicates less likelihood.

#### Interpretation:

#### - Among the 18- 44 years audience, the most popular categories of gifts are Clothing and Jewelry. For 45+ audience, the most popular category of gifts is Greeting Cards.

#### - We may observe the overall trend (pink line) that the percent of people who buy gifts decreases over time.

## Affinity Index vs Gender

```{r}
gifts_gender_affinity_plot <- gifts_gender_ext_wide %>% 
  select(Gender, starts_with("Affinity_")) %>% 
  pivot_longer(
    cols= starts_with("Affinity_"),
    names_to = "Category",
    values_to = "AffinityIndex"
  ) %>% 
  mutate(
    Category = gsub("Affinity_", "", Category),
    Gender = factor(Gender, levels=c("Men", "Women"))
  )

ggplot(gifts_gender_affinity_plot, aes(x=Category, y=AffinityIndex, fill=Gender)) + 
  geom_col(position="dodge") + 
  geom_hline(yintercept = 100, linetype="dashed", color="grey50", linewidth=0.8) +   theme_minimal() + 
  labs(
    title="Affinity Index vs Gift Categories (By Gender)",
    y = "Affinity Index",
    x = "Gift Category"
  ) + 
  scale_fill_manual(values=c("Men"="steelblue", "Women" = "lightcoral")) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### The above plot visualizes the affinity indexes for each category of gifts and gender. We decided to include this plot as it shows the most popular categories of gifts. An Affinity Index above 100 indicates that a gender is more likely to buy that category of gifts compared to the average (100), while below 100 indicates less likelihood.

#### Interpretation:

#### - On average, women are more likely to buy gifts with the exception of Flowers and Jewelry.

### Change of Total Spending over time

```{r}
inflation_data <- tibble(
  Year = c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025),
  Amount = c(1, 1.03, 1.05, 1.07, 1.09, 1.09, 1.1, 1.12, 1.15, 1.17, 1.19, 1.24, 1.34, 1.4, 1.44, 1.47),
  InflationRate = c(0.02, 0.03, 0.02, 0.01, 0.02, 0, 0.01, 0.02, 0.02, 0.02, 0.01, 0.05, 0.08, 0.04, 0.03, 0.02)
) %>%
  rename(CPI_Index = Amount)

historical_spending_w_inflation <- historical_spending_ext %>%
  left_join(inflation_data %>% select(Year, CPI_Index), by = "Year")

base_year_cpi_index <- inflation_data %>% filter(Year == 2022) %>% pull(CPI_Index)

historical_spending_adjusted <- historical_spending_w_inflation %>%
  mutate(
    Real_AnnualSpending_tot = AnnualSpending_tot * (base_year_cpi_index / CPI_Index)
  ) %>%
  select(Year, AnnualSpending_tot, Real_AnnualSpending_tot) %>%
  pivot_longer(
    cols = c(AnnualSpending_tot, Real_AnnualSpending_tot),
    names_to = "SpendingType",
    values_to = "SpendingValue"
  )

ggplot(historical_spending_adjusted, aes(x = Year, y = SpendingValue, color = SpendingType)) +
  geom_line(size = 1.2) +
  geom_point(size = 3, shape = 19) +
  scale_x_continuous(breaks = unique(historical_spending_adjusted$Year)) +
  scale_color_manual(
    values = c("AnnualSpending_tot" = "purple", "Real_AnnualSpending_tot" = "darkgreen"),
    labels = c("AnnualSpending_tot" = "Nominal Spending", "Real_AnnualSpending_tot" = paste0("Real Spending (2022 $)"))
  ) +
  theme_minimal() +
  labs(
    title = "Total Spending Over Time",
    y = "Total Annual Spending ($)",
    x = "Year",
    color = "Spending Type"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

#### The above plot represents the change of spending over time. We took into account inflation of dollar and added an adjusted line (green) to visualize the actual spending volume change. The inflation data has been take from the open source: <https://www.in2013dollars.com/us/inflation/2010?endYear=2022&amount=1>

#### Interpretation:

#### - Both Nominal and Real Spending lines indicate that there's an overall increase in the actual purchasing power of money spent over time.

### Average Spending per Gift category

```{r}
historical_spending_ext_long <- historical_spending_ext %>% 
  select(1, 4:(ncol(historical_spending_ext) - 3)) %>% 
  pivot_longer(
    cols = -Year,
    names_to = "Category",
    values_to = "AverageSpending"
  )

historical_spending_ext_long
ggplot(historical_spending_ext_long, aes(x=Year, y=AverageSpending, fill=Category)) + 
  geom_area(alpha=0.8, position = "stack") + 
  scale_x_continuous(breaks=unique(historical_spending_ext_long$Year)) + 
  theme_minimal() + 
  labs(
    title = "Average Spending per Gift category over Time",
    y = "Average Spending",
    x = "Year",
    fill = "Gift Category"
  ) + scale_fill_brewer(palette = "Paired") + 
  theme(plot.title = element_text(hjust=0.5))
```

#### Stacked area plot appears to be a perfect tool to visualize the proportion of spending on each category of gifts and the overall annual change.

#### Interpretation:

#### - Jewelry and Evening Out represent the largest portions of total spending over the observed period

#### - Spending on all categories showed limited growth, even experiencing declines, during the 2015-2019 period.

## Distribution of Spending Across Each Gift Category Over All Years

```{r}
historical_spending_long_categories <- historical_spending_ext %>%
  select(Year, Candy:GiftCards) %>%
  pivot_longer(cols = -Year, names_to = "Category", values_to = "Spending")

ggplot(historical_spending_long_categories, aes(x = Category, y = Spending, fill = Category)) +
  geom_boxplot() +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = "Distribution of Spending Across Gift Categories Over All Years",
    x = "Gift Category",
    y = "Spending ($)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

#### This box plot displays the range, median, and variability of actual dollar spending for each specific gift category over all years, serving to highlight which categories consistently involve higher or lower expenditures and their spread.

## Gift Category Spending Distribution by Age

```{r}

gifts_age_percentages_long_categories <- gifts_age_ext %>%
  select(Age, Age_cat, Candy:GiftCards) %>% 
  pivot_longer(
    cols = Candy:GiftCards, 
    names_to = "Category",
    values_to = "PercentageOfPeople"
  ) %>%

  mutate(Age = factor(Age, levels = unique(gifts_age_ext$Age[order(gifts_age_ext$Age_cat)])))

ggplot(gifts_age_percentages_long_categories, aes(x = Category, y = PercentageOfPeople, fill = Category)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "Distribution of Percentage of People Spending Across Gift Categories by Age Group",
    x = "Gift Category",
    y = "Percentage of People (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") 
```

#### This box plot visualizes the statistical distribution of the percentage of people who purchase different gift categories, allowing for a clear comparison of varying buying prevalences across those categories within specific age groups.

## Percentage of Celebrating by Age Group

```{r}
ggplot(gifts_age_ext_wide, aes(x=Age, y=SpendingCelebrating)) + 
  geom_col(fill="steelblue") +
  geom_text(aes(label=paste0(round(SpendingCelebrating, 1), "%")),
            vjust= -0.5,
            color="black",
            size=3) + 
  theme_minimal() +
  labs(
    title = "Percentage of People Celebrating by Age",
    y = "Percent of Celebrating",
    x = "Age Group"
  ) + 
  theme(axis.title.x = element_text(angle = 45, hjust=1))

```

#### This plot compares how many individuals from every age group are celebrating Saint Valentine's Day

#### Interpretation:

#### - The 18-24 appears to have the highest percentages of people celebrating, there's a strictly inversed dependency with the age

```{r}
library(dplyr)

# Add new columns:
# - Total: total spending across all gift categories
# - Avg_Gift: average spending per gift category
gifts_age_ext <- gifts_age %>%
  mutate(
    Total = Candy + Flowers + Jewelry + GreetingCards + EveningOut + Clothing + GiftCards,
    Avg_Gift = Total / 7  # Divide by number of gift types
  )
```

### Description:

#### This code creates a new version of the gifts_age dataset with additional columns showing total spending on gifts and average gift spending per age group.

```{r}
library(ggplot2)

ggplot(gifts_age_ext, aes(x = Age, y = Flowers, group = 1)) +
  geom_line(color = "#E63946", size = 1.2) +
  geom_point(color = "#E63946", size = 3) +
  point
  labs(
    title = "Average Spending on Flowers by Age Group",
    x = "Age Group",
    y = "Average Spending (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```

### Description:

#### This line chart visualizes how much different age groups spend on flowers for Valentine's Day. It's great for spotting age-based trends.

```{r}
library(tidyr)

# Reshape the data into long format for easier plotting
gifts_long <- gifts_age_ext %>%
  select(Age, Candy, Flowers, Jewelry, GreetingCards, EveningOut, Clothing, GiftCards) %>%
  pivot_longer(-Age, names_to = "GiftType", values_to = "Spending")

# Create a stacked bar chart by gift type and age group
ggplot(gifts_long, aes(x = Age, y = Spending, fill = GiftType)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Valentine's Day Spending by Age Group and Gift Type",
    x = "Age Group",
    y = "Spending (%)",
    fill = "Gift Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Description:

#### This stacked bar chart shows the distribution of gift spending types by age group. Each bar is divided by gift type, showing which categories dominate.

```{r}
# Plot average spending per age group
ggplot(gifts_age_ext, aes(x = Age, y = Avg_Gift)) +
  geom_col(fill = "#457B9D") +  # Blue bars
  labs(
    title = "Average Gift Spending by Age Group",
    x = "Age Group",
    y = "Average Spending (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Description:

#### This bar chart summarizes the average total spending per person in each age group across all gift categories.

Save processed data

```{r}
write.csv(gifts_gender_ext_wide, 'gifts_gender_processed.csv')
```

```{r}
write.csv(gifts_age_ext_wide, 'gifts_age_processed.csv')
```

```{r}
write.csv(historical_spending_ext, 'historical_spending_processed.csv')
```

# Part II

```{r}
libs <- c("tidyverse", "tidymodels", "vip", "rpart.plot", "kknn", "ggplot2")

installed_libs <- libs %in% rownames(installed.packages())

if (any(installed_libs == FALSE)) {
  install.packages(libs[!installed_libs])
}

# Load libraries
library(tidyverse)
library(tidymodels)
library(vip)          # For variable importance plots
library(rpart.plot)
library(kknn)
library(ggplot2)
```

```{r}
gift_gender_processed <- read_csv('gifts_gender_processed.csv')
gifts_age_processed <- read_csv('gifts_age_processed.csv')
historical_spending_processed <- read_csv('historical_spending_processed.csv')
```

### Decision Tree

```{r}
ga_recipe_tree <- recipe(
  EveningOut ~ ., 
  data = gifts_age_processed 
) %>%
  step_rm(Age, HighestSpending) %>% 
  step_YeoJohnson(all_numeric_predictors(), -all_outcomes()) %>% 
  step_normalize(all_numeric_predictors(), -all_outcomes()) 
prep(ga_recipe_tree) %>%
  bake(new_data = head(gifts_age_processed))
```

```{r}

tree_spec_reg <- decision_tree(
  cost_complexity = 0.01, 
  tree_depth = 3,         
  min_n = 2               
) %>%
  set_engine("rpart") %>%
  set_mode("regression")

print(tree_spec_reg)

tree_workflow_reg <- workflow() %>%
  add_model(tree_spec_reg) %>%
  add_recipe(ga_recipe_tree)

print(tree_workflow_reg)

fitted_tree_reg <- fit(tree_workflow_reg, data = gifts_age_processed)

print(fitted_tree_reg)

final_tree_rpart_fit_simple <- extract_fit_engine(fitted_tree_reg)

```

```{r}
rpart.plot(
  final_tree_rpart_fit_simple,
  roundint = FALSE,
  box.palette = "BuGn",
  extra = 1,
  tweak = 1.2,
  main = "Decision Tree Regression"
)
```

```{r}
vip(fitted_tree_reg) +
  labs(title = "Decision Tree Regression Variable Importance")

tree_predictions <- predict(fitted_tree_reg, new_data = gifts_age_processed) %>%
  bind_cols(gifts_age_processed %>% select(EveningOut)) 

metrics_simple <- tree_predictions %>%
  metrics(EveningOut, .pred)

print("Metrics on the full 6-record dataset:")
print(metrics_simple)
```

```{r}
ggplot(tree_predictions, aes(x = EveningOut, y = .pred)) +
  geom_point(size = 3) +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed', color = 'red') +
  labs(title = 'Decision Tree Regression: Predicted vs. Actual EveningOut Spending (Full Dataset)',
       subtitle = 'Model fitted on all 6 records - results prone to overfitting',
       x = 'Actual EveningOut Spending',
       y = 'Predicted EveningOut Spending') +
  theme_minimal()
```

### Linear Regression

```{r}
historical_spending_lm <- historical_spending_processed %>% 
  select(-AnnualSpending_tot, -HighestSpending)

lr_model <- lm(Jewelry ~ Year, data = historical_spending_lm)

# Create future years for prediction (5 years ahead)
future_years <- tibble(Year = 2024:2028)

# Predict jewelry spending for future years
future_years <- future_years %>%
  mutate(Predicted_Jewelry = predict(lr_model, newdata = future_years))

# Combine historical and predicted data for plotting
plot_data <- historical_spending_lm %>%
  select(Year, Jewelry) %>%
  mutate(Type = "Actual") %>%
  bind_rows(
    future_years %>%
      rename(Jewelry = Predicted_Jewelry) %>%
      mutate(Type = "Predicted")
  )

# Plot the data
ggplot(plot_data, aes(x = Year, y = Jewelry, color = Type)) +
  geom_point(size = 3) +
  geom_line() +
  labs(title = "Jewelry Spending Forecast",
       y = "Jewelry Spending (USD)",
       x = "Year") +
  theme_minimal()
```

```{r}
summary(lr_model)
```

### KNN Regression

```{r}
historical_spending_knn <- historical_spending_processed %>%
  select(-Year, -AnnualSpending_tot, -HighestSpending, -SpendingChange)
historical_spending_knn
```

```{r}
set.seed(314)

hist_split <- initial_split(historical_spending_knn, prop = 0.9)

hist_training <- training(hist_split)
hist_test <- testing(hist_split)
cat('Historical spending training set dimensions:', dim(hist_training), '\n')
cat('Historical spending testing set dimensions:', dim(hist_test), '\n')
```

```{r}
hist_recipe <- recipe(
  Jewelry ~ ., 
  data = hist_training 
) %>%
  step_YeoJohnson(all_numeric_predictors(), -all_outcomes()) %>% 
  step_normalize(all_numeric_predictors(), -all_outcomes()) 
```

```{r}
hist_recipe_prepped <- prep(hist_recipe, training = hist_training)
baked_hist_training <- bake(hist_recipe_prepped, new_data = hist_training)
cat('First few rows of baked historical spending training data:\n')
glimpse(baked_hist_training)
```

```{r}
knn_model_spec_reg <- nearest_neighbor(neighbors = tune()) %>% 
  set_engine('kknn') %>% 
  set_mode('regression')

knn_model_spec_reg
```

```{r}
knn_wf_reg <- workflow() %>%
  add_model(knn_model_spec_reg) %>%
  add_recipe(hist_recipe)

knn_wf_reg
```

```{r}
k_grid_reg <- tibble(neighbors = c(1, 2))
k_grid_reg
```

```{r}
set.seed(314)
hist_folds <- vfold_cv(hist_training, v = 2)
knn_tuning_reg <- tune_grid(
  knn_wf_reg,
  resamples = hist_folds,
  grid = k_grid_reg,
  metrics = metric_set(rmse, rsq, mae)
)

knn_tuning_reg
```

```{r}
autoplot(knn_tuning_reg) +
  labs(title = 'KNN Hyperparameter Tuning (Historical Spending Data)',
       subtitle = 'Performance vs. Number of Neighbors (k)') +
  theme_light()

```

```{r}
knn_tuning_reg %>%
  show_best(metric = 'rmse', n = 3)

best_k_reg <- knn_tuning_reg %>%
  select_best(metric = 'rmse')

print('Best k for regression:')
print(best_k_reg)
```

```{r}
final_knn_wf_reg <- knn_wf_reg %>%
  finalize_workflow(best_k_reg)

final_knn_wf_reg
```

 

```{r}
last_fit_knn_reg <- final_knn_wf_reg %>%
  last_fit(split = hist_split)
```

```{r}
last_fit_knn_reg %>%
  collect_metrics()
```

```{r}
knn_test_predictions_reg <- last_fit_knn_reg %>%
  collect_predictions()
```

```{r}
ggplot(knn_test_predictions_reg, aes(x = Jewelry, y = .pred)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed', color = 'red') +
  labs(title = 'KNN Regression: Predicted vs. Actual Jewelry Spending (Test Data)',
       x = 'Actual Jewerly Spending',
       y = 'Predicted Jewerly Spending') +
  theme_minimal()
```

## Summary

### 1. Decision Tree for Regression for gifts_age

-   'EveningOut' was set as a target variable, 'HighestSpending' feature was removed as it was derived based on comparison of other features including the target variable, so there's no sense to keep it. Prediction of 'EveningOut' variable would be helpful for entertainment places such as cinemas, restaurants, pubs, etc.

-   Interpretation of results: the Decision Tree graph indicates that the target variable primarily relies on the Age Group (set as ...1 on the graph)

-   Model's fit assessment: models fit was assessed with RMSE, Rsq and MAE metrics. Additionally, visualizations such as 'Predicted vs Actual' plot and Variable Importance plots were used.

-   Notes: gifts_age dataset is extremely small, that's why the decision tree was trained and tested on the whole dataset. Otherwise, the model would consist of a single split.

### 2. Linear Regression for historical spending

-   'Jewerly' spending was set a target variable, we used 'Year' as a single dependent feature to detect the time-based trend. Forecasting the jewerly spending would be helpful for the jewerly stores.

-   Interpretation of the results: p-value indicates that 'Year' is significant feature.

-   Model's fit assessment: model's fit was assessed with multiple and adjusted R2, F-statistics, residual standard error as well as visual inspection of a plot. Our linear regression model explains approximately 56.6 % of the variance in Jewerly based on the Year.

### 3. KNN for historical spending

-   This model addresses the problem from Model 2 but with a different approach. The objective is to predict 'Jewerly' but with usage of all variables except the 'Year', 'AnnualSpending_tot', 'HighestSpending', 'SpendingChange'.

-   Interpretation of the results: there's no simple variable importance or p-statistics. KNN computes the average between K nearest neighbors

-   Model's fit assessment: model's fit with RMSE and Rsq metrics as well as 'Predicted vs Actual' plot.

### Models comparison

In this project, we have examined 3 different models, the Linear Regression and Decision Tree showed high interpretability. We approached prediction of 'Jewerly' spending with Linear Regression and KNN Regression with different techniques. KNN Regression has higher R2 however, it would be unfair to compare that metric directly as the dataset is extremely small and KNN is definetly overfitting.

In conclusion, our Linear Regression model stands out for its high interpretability and clear statistical significance, effectively demonstrating a linear trend in 'Jewelry' spending over 'Year'. The Decision Tree, while also highly interpretable, faces significant challenges due to the extremely small dataset used for 'EveningOut' prediction, making its error metrics less reliable. The KNN model offers a different approach to predicting 'Jewelry' spending, focusing on similarity, but sacrifices direct interpretability for potentially complex non-linear relationships.

# Project Summary

# Steps taken

## Part I EDA

### Data Loading and Initial Exploration (EDA):

-   Three datasets were loaded: `historical_spending.csv`, `gifts_age.csv`, and `gifts_gender.csv`.

-   Initial data understanding was performed using `glimpse()` and `skim()` to inspect structure, data types, and summary statistics, including checks for missing values. None of datasets contains missing values

### Data Extension / Feature Engineering: New, insightful features were computed for each dataset to enrich the analysis:

-   `historical_spending` dataset:

    -   `AnnualSpending_tot`: Calculated as the total spending across all gift categories for each year, providing an overall spending trend.

    -   `SpendingChange`: Computed the year-over-year percentage change in `PerPerson` spending to observe spending growth or decline.

    -   `HighestSpending`: Identified the gift category with the highest spending each year, highlighting dominant categories.

-   `gifts_age` dataset:

    -   `Age_cat`: Converted categorical age ranges into numerical categories (1-6) for easier processing in models.

    -   `HighestSpending`: Determined the gift category with the highest spending for each age group.

    -   `Affinity_` indices: Calculated an "Affinity Index" for each gift category per age group. An index \> 100 indicates that an age group is more likely to purchase a product compared to the overall average, providing insights into group-specific preferences.

-   `gifts_gender` dataset:

    -   `HighestSpending`: Identified the gift category with the highest spending for each gender.

    -   `Affinity_` indices: Computed "Affinity Index" for each gift category per gender, similar to the age groups, to understand gender-specific purchasing likelihoods.

### Data Visualization (EDA Plots)

A variety of plots were generated to visualize trends, distributions, and relationships within the data:

-   Affinity Index vs Age Group: Showed popular gift categories for different age ranges (e.g., Clothing and Jewelry for 18-44, Greeting Cards for 45+), and an overall trend of decreasing gift-buying percentage with age.

-   Affinity Index vs Gender: Illustrated gender-based spending likelihoods, noting women's higher likelihood to buy gifts except Flowers and Jewelry.

-   Change of Total Spending over Time: Compared nominal and real (inflation-adjusted) total spending, indicating an overall increase in purchasing power over time.

-   Average Spending per Gift Category over Time (Stacked Area Plot): Visualized the proportion of spending on each category annually, showing Jewelry and Evening Out as largest portions and a limited growth from 2015-2019.

-   Distribution of Spending Across Each Gift Category: Box plots displayed the range, median, and variability of spending for each category.

-   Gift Category Spending Distribution by Age: Box plots to compare buying prevalences across categories within age groups.

-   Percentage of Celebrating by Age Group: showed that the 18-24 age group had the highest percentage of people celebrating Valentine's Day, with a decreasing trend as age increases.

-   Average Spending on Flowers by Age Group: showed that the 18-24 age group had the highest percentage of people celebrating Valentine's Day, with a decreasing trend as age increases.

-   Valentine's Day Spending by Age Group and Gift Type: This stacked bar chart shows the distribution of gift spending types by age group. Each bar represents an age group and is divided by different gift types, illustrating which categories dominate within each age segment. In younger age groups like 18-24, Candy and Evening Out appear to constitute larger portions, while in older groups, other categories might become more prominent such as Greeting cards.

-   Average Gift Spending by Age Groups: This bar chart summarizes the average total spending per person in each age group across all gift categories.There's a general trend of decreasing average gift spending as age increases, which aligns with the "Percentage of Celebrating by Age Group" plot.

## Part II Modelling

During that phase, three regression models were developed to predict spending, each with a different objective and approach. Including:

-   **Decision Tree for Regression:**

    Objective: Predict `EveningOut` spending using the gifts_age_processed dataset.

    Preprocessing: `Age` and `HighestSpending` were removed, and remaining numeric predictors underwent Yeo-Johnson transformation and normalization.

    Model Specification: A decision_tree model with cost_complexity = 0.01, tree_depth = 3, and min_n = 2 was defined using the rpart engine.

    Evaluation: The model was fitted and evaluated on the entire 6-record gifts_age_processed dataset, using RMSE, R-squared, and MAE. Variable importance and the tree structure were visualized. The model faces limitations due to the dataset's size.

-   **Linear Regression:**

    Objective: Forecast Jewelry spending based on Year using historical_spending_lm to detect time-based trends.

    Model: A simple linear model lm(Jewelry \~ Year) was fitted.

    Evaluation: The summary() of the model provided statistical insights, including R-squared (0.5655), residual standard error (4.26), and p-values (showing Year as a significant predictor). A forecast plot visualized historical trends and future predictions.

-   **KNN Regression:**

Objective: Predict Jewelry spending using other spending categories and Age_cat from historical_spending_knn, offering an alternative approach to the linear model.

Preprocessing: Numeric predictors were transformed using Yeo-Johnson and normalized.

Data Split: The dataset was split into 90% training and 10% testing sets using initial_split().

Hyperparameter Tuning: A k grid of 1 and 2 neighbors was tuned using 2-fold cross-validation (vfold_cv) on the training data, optimizing for RMSE.

Evaluation: The final workflow was fitted to the training data and evaluated on the test set using last_fit(), collecting RMSE, R-squared, and MAE. A plot of predicted vs. actual values on the test data was generated.

# Conclusions and Reflections

## What went well-defined and what was difficult?

### Well

The project successfully demonstrated various data manipulation and visualization techniques, effectively telling a story through plots. The application of different models (tree, linear, KNN) was also well-executed. Feature engineering, such as creating `HighestSpending` and `Affinity_` indices, provided valuable new insights.

### Challenges

A significant challenge was the extremely small size of the provided datasets (13, 6 and 2 records), which severely limited the robustness of the Decision Tree model's evaluation and made proper train/test splitting for generalizable performance assessment difficult. Directly comparing the Linear Regression's "residual standard error" (a training-based metric) with KNN's test set metrics was statistically inappropriate without re-evaluating the linear model on a test set.

## What could have been done better?

Consistent Evaluation: All models, especially if intended for performance comparison, should have been evaluated on a consistent, independent test set. This would involve creating a single train/test split at the beginning for all models predicting `Jewelry` or `EveningOut` that share data. However, this would require larger datasets. What was the most challenging, and what was the most interesting?

## What was the most challenging, interesting?

### Most Challenging

-   Dealing with the extremely small datasets and the implications for model reliability and evaluation was likely the most challenging aspect.
-   The issue with the KNN's R-squared also presented a significant challenge.

### Most Interesting:

-   The feature engineering, particularly the creation of `Affinity_` Indices and `HighestSpending`, was very interesting as it provided a clear, interpretable measure of spending likelihood by age and gender.
-   The diverse set of visualizations, from stacked area plots to box plots and time series forecasts, effectively communicated hidden data patterns.

## What was the most useful?

-   The `Affinity_` Index feature was very useful for understanding consumer preferences by demographic.

-   The tidymodels framework provided a consistent and robust way to build and evaluate the Decision Tree and KNN models, promoting good machine learning practices (like preprocessing pipelines and hyperparameter tuning).

-   The Linear Regression model was particularly useful for its high interpretability and clear statistical significance in demonstrating a trend in `Jewelry` spending over time.

-   The various visualizations were invaluable for initial data exploration and communicating insights.

## What would you do differently if you were starting the project from scratch?

-   Extend the datasets: The obtained datasets are too small to build meaningful models. We could try extending the datasets by generating more data (i.e. with AI) or searching for larger datasets.

-   Unified Data Splitting: Implement a single, unified train/test split for all models that predict the same target variable (e.g., Jewelry) to ensure fair comparison of their generalization performance.

## What would you do differently if you had more time?

-   Collect More Data or Extend the Datasets: If feasible, acquire or simulate more data, to enable more robust modeling.

-   Broader Hyperparameter Tuning: Conduct more extensive hyperparameter tuning for the KNN model (and potentially the Decision Tree) with a wider range of k values and more cross-validation folds after obtaining larger datasets.

-   Advanced Models: Explore more complex or ensemble modeling techniques (e.g., Random Forests, Gradient Boosting Machines) that might capture more intricate relationships.

## What would you do differently if you had more knowledge/resources?

-   Collect More Data or Extend the Datasets: Acquire or simulate more data, to enable more robust modeling. Alternatively, we could select different datasets.
